---
title: "Module 2 - Question 4"
author: "Amadou Beydi Coulibaly"
date: "2026-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
```

# Task 4: Exploring the hospital length-of-stay dataset

Before we can build a Bayesian regression for length of stay (Task 5), we need
to understand the dataset: what the target variable looks like, which
predictors are available, and any data-quality quirks that could bite us later.

## 1. Loading the data and overview

```{r load}
mimic <- read_csv("mimic3d1.csv")
dim(mimic)
glimpse(mimic)
```

The dataset contains **58,976 hospital admissions and 28 variables**. Each row
is one admission (`hadm_id`), with demographics (`gender`, `age`, `insurance`,
`ethnicity`, ...), admission details (`admit_type`, `admit_location`,
`AdmitDiagnosis`), a set of `Num*` columns measuring how intensively the
hospital interacted with the patient (labs, prescriptions, notes, transfers,
...), an in-hospital mortality flag (`ExpiredHospital`), and our target:
**`LOSdays`**, the length of stay in days.

## 2. The target variable: LOSdays

```{r los-summary}
mimic %>% summarise(
  mean     = mean(LOSdays),  median = median(LOSdays),
  sd       = sd(LOSdays),    min    = min(LOSdays),  max = max(LOSdays),
  skew     = e1071::skewness(LOSdays),
  p_over_2 = mean(LOSdays > 2)
)
```

Three things stand out:

* The distribution is **heavily right-skewed** (skewness ≈ 4.6): the median
  stay is about 6.5 days, but the mean is pulled up to 10.1 days by a long
  tail that reaches **295 days**.
* **89.7% of admissions last longer than 2 days** — a useful empirical
  baseline for the Task 5 question about the probability of "definitely
  missing work".
* There are **128 admissions with `LOSdays = 0`**, which matters whenever we
  take logs.

```{r los-hist, fig.width=9, fig.height=4.5}
los_mean   <- mean(mimic$LOSdays)
los_median <- median(mimic$LOSdays)

ggplot(mimic, aes(LOSdays)) +
  geom_histogram(binwidth = 1, fill = "#1F78B4", alpha = 0.85) +
  geom_vline(xintercept = los_median, color = "#8F001A", linewidth = 1) +
  geom_vline(xintercept = los_mean,   color = "#8F001A",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = los_median, y = Inf, vjust = 2, hjust = -0.1,
           label = sprintf("Median = %.1f", los_median), color = "#8F001A", size = 3.5) +
  annotate("text", x = los_mean, y = Inf, vjust = 4, hjust = -0.1,
           label = sprintf("Mean = %.1f", los_mean), color = "#8F001A", size = 3.5) +
  coord_cartesian(xlim = c(0, 60)) +   # zoom; tail extends to ~295 days
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16, color = "#8F001A"),
        plot.subtitle = element_text(size = 12, color = "grey40", margin = margin(b = 10))) +
  labs(title = "Length of Stay is Heavily Right-Skewed",
       subtitle = "Most stays are short, but a long tail pulls the mean above the median",
       x = "Length of stay (days)", y = "Number of admissions")
ggsave("task4_los_hist.pdf", width = 9, height = 4.5, units = "in")
```

A skewed, strictly-positive outcome is a poor fit for a plain normal
likelihood. The standard remedy is to model the **log** of the outcome, so we
check whether that helps:

```{r los-log, fig.width=9, fig.height=4.5}
mimic %>%
  filter(LOSdays > 0) %>%                 # log undefined at 0 (drops 128 rows)
  ggplot(aes(log(LOSdays))) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "#1F78B4", alpha = 0.85) +
  stat_function(fun = dnorm, color = "#8F001A", linewidth = 1,
                args = list(mean = mean(log(mimic$LOSdays[mimic$LOSdays > 0])),
                            sd   = sd(log(mimic$LOSdays[mimic$LOSdays > 0])))) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16, color = "#8F001A"),
        plot.subtitle = element_text(size = 12, color = "grey40", margin = margin(b = 10))) +
  labs(title = "On the Log Scale, Length of Stay is Roughly Normal",
       subtitle = "Skewness falls from 4.6 to -0.5; a normal model is reasonable on log-days",
       x = "log(Length of stay)", y = "Density")
ggsave("task4_los_log.pdf", width = 9, height = 4.5, units = "in")
```

On the log scale the skewness drops from 4.6 to about −0.5 and the histogram
tracks a normal curve closely. **This justifies using normal priors and a
normal likelihood on `log(LOSdays)` in Task 5**, then transforming back to
days to answer probability questions like P(LOS > 2).

## 3. How LOS varies across admissions

```{r los-by-type, fig.width=9, fig.height=4.5}
mimic %>%
  ggplot(aes(x = reorder(admit_type, LOSdays, median), y = LOSdays)) +
  geom_boxplot(fill = "#1F78B4", alpha = 0.6, outlier.alpha = 0.1) +
  coord_cartesian(ylim = c(0, 30)) +      # zoom past the outliers
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16, color = "#8F001A")) +
  labs(title = "Length of Stay by Admission Type",
       x = NULL, y = "Length of stay (days)")
ggsave("task4_los_by_type.pdf", width = 9, height = 4.5, units = "in")

mimic %>%
  group_by(admit_type) %>%
  summarise(n = n(), median = median(LOSdays), mean = mean(LOSdays))
```

Emergency admissions dominate the data (about 71%). Urgent admissions stay
longest (median 8.1 days). Newborns are an interesting subgroup: they have the
**shortest median (4.1 days) but a high mean (11.5 days)** — most newborns go
home quickly while a minority of NICU cases stay for weeks, another sign of
the heavy right tail.

```{r los-by-demo}
mimic %>% group_by(gender) %>%
  summarise(n = n(), median = median(LOSdays), mean = mean(LOSdays))

summary(mimic$age)
```

Gender makes essentially no difference (median 6.4 vs 6.5 days). Age runs
from 0 to 89 (capped at 89 for de-identification, and newborns all have
age 0), with a typical patient around 59 years old.

## 4. Numeric predictors: a caveat about the `Num*` columns

```{r correlations}
mimic %>%
  select(where(is.numeric), -hadm_id, -LOSgroupNum) %>%
  cor() %>%
  as_tibble(rownames = "variable") %>%
  select(variable, cor_with_LOS = LOSdays) %>%
  filter(variable != "LOSdays") %>%
  arrange(desc(abs(cor_with_LOS)))
```

```{r rates-check}
mimic %>% select(LOSdays, NumRx, NumLabs, NumNotes, NumCallouts) %>% head()
```

Two findings here:

* All linear correlations with `LOSdays` are **weak** (|r| ≤ 0.21), so no
  single predictor will explain LOS by itself.
* The `Num*` columns contain **non-integer values** (e.g. 14.91
  prescriptions), so they are **per-day rates, not raw counts** — the totals
  have already been divided by the length of stay. That explains why several
  of them correlate *negatively* with LOS: short stays tend to be intense
  (many interactions per day), long stays less so. We should keep this in
  mind when interpreting regression coefficients in Task 5.

## 5. Data quality

```{r missingness}
mimic %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

# AdmitProcedure hides its missing values as the literal string "na"
mimic %>% summarise(admitproc_na = sum(AdmitProcedure == "na"))
```

Genuine `NA`s are confined to `marital_status` (10,128), `religion` (458) and
`AdmitDiagnosis` (25). In addition, `AdmitProcedure` encodes missingness as
the literal string `"na"` in 8,032 rows — a trap if we ever filter with
`is.na()` alone.

## Key takeaways for Task 5

1. `LOSdays` is strictly positive and heavily right-skewed; **model it on the
   log scale**, where it is approximately normal — consistent with the hint
   to use normal priors.
2. The 128 zero-day stays must be handled (dropped or offset) before taking
   logs.
3. **89.7% of stays already exceed 2 days**, so the posterior probability of
   staying longer than 2 days should come out high.
4. The `Num*` predictors are per-day rates with weak linear correlations to
   LOS — useful covariates, but their signs need careful interpretation.
5. Missingness is mild and concentrated in demographic fields; watch out for
   the `"na"` strings in `AdmitProcedure`.
